Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-12 EBDVF-2024/paper.qmd
edits

---
title: Raising the role of Vocabulary Hubs for Semantic data interoperability in Dataspaces

@@

author:
  - name: Martin Kaltenböck
    email: martin.kaltenboeck@semantic-web.com
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    affiliation: 
     name: The Semantic Web Company
     country: Austria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliation: 
      name: Ontotext
      country: Bulgaria
acm-metadata:
  keywords:
    - data spaces
    - semantic interoperability
    - semantic technologies
    - ontologies
    - vocabulary hub
    - oil and gas
    - renewable energy
    - refineries
    - windfarms
  final: true
  # acmart-options: sigconf, review
  # copyright-year: 2024
  # acm-year: 2024
  copyright: acmcopyright
  # doi: 
  # conference-acronym: 
  # conference-name: 
  # conference-date: 
  # conference-location: 
  # isbn:

@@

# Introduction
Consolidating wind farm data in data spaces for training predictive maintenance ML - Use Case Description

# Initial Situation
In the UNDERPIN project, we face the challenge of training ML for predictive maintenance of wind farms. This scenario benefits from large amounts of training data. Such data can be provided by different partners who collaborate on this ML approach. We support this scenario by implementing a data space for wind farm data. Data spaces enable data sharing between trusted participants while keeping data sovereignty. However, for consolidating different data sets from participants, we need to establish a clear semantics and mapping for the data.

# Approach and IT-Solution
To tackle this problem, we introduce Semantic Web standards and technologies to model and process this data. Vocabularies and ontologies, specifically expressed using OWL, can represent such data with a clear semantics based on standards. Furthermore, we introduce GraphDB and PoolParty as two software components which implement these standards to process the data.

## IDS Vocabulary Hub
IDS defines the architecture of data spaces in the reference architecture model IDS-RAM. One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services. These vocabularies form a common language to enable semantic interoperability among the data space participants. The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of a standardized semantic descriptions. It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies and thereby provides semantic  interoperability in a standardized and machine-readable way.

## GraphDB Graph Database
tbd

## PoolParty Semenatic Suite
PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.

@@

PoolParty is well suited to form the basis of a vocabulary management system within a data space, effectively implementing not only the Vocabulary Hub role, but furthermore providing additional services which improve the expressivity of common vocabularies and automate semantic interoperability by intelligent data analysis.

@@

Both GraphDb and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL. They therefore seamlessly integrate with each other, but also with other services and components in data spaces which support these machine-readable standards for semantic interoperability.

The Wind Farm Ontology [@ikeresnaola-gonzalezWindFarmOntology2021] ...

# Business Value and Benefits of the Semantic Solution
When consolidating the data, we need to provide a method to automatically map and process it so this can be done efficiently on large data sets and/or regular updates in data spaces. The benefit provided should support the ML training in such a way that the consolidated data set is exposed as uniform training data set.

## the Semantic Layer Approach
The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in data spaces by provisioning common vocabularies. We aim to go one step further and envision a Semantic Layer for data spaces, which provide vocabulary-based services for advanced and automated semantic descriptions of metadata and data. The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.

In the following, we discuss i) the services provided for the Vocabulary Hub role and ii) additional services as extensions to the Vocabulary Hub, which implement towards a Semantic Layer for data spaces.

Vocabulary Hub Services:
* RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
* scalable storage of vocabularies: cluster implementation for higher throughput and a fault tolerant architecture.
* knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. Vocabularies can be created and edited based on standards like SKOS and OWL, but also existing vocabularies can be reused and combined to fit specific needs of data spaces.
* SPARQL service for semantic lookups: dereference of URIs describing data asset to retrieve a semantic description of the entity from the graph database.

Semantic Layer Services:
* we support both unstructured and structured data assets
+ unstructured data: semantic annotation services for documents based on concept tagging. PoolParty’s KG-based NLP  primarily uses SKOS taxonomies to identify concepts in documents, but can be extended with OWL ontology elements (classes, properties) for increased expressivity.
+ structured data: we support methods for structured data which can be decided based on the actual content in place, ranging from OWL ontologies for OBDA, linking SKOS concepts and
- with OBDA for data spaces we support an ontology-based method which is well suited for data retrieval while keeping the original source in place. Ontologies for OBDA are intended to be governed via the Semantic Layer and can consolidate data assets of various formats using a precise mapping. Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
- R2RML mapping is intended for data transformation processes within a data space, where the original source is used as a basis to create an RDF representation.
- SKOS concepts can also be used for structured data in several ways. Concepts can be directly added to RDF data via semantic links (properties from an ontology), they can be used to annotate specific literals of an RDF graph or they can be used to associate documents with (specific parts of) an RDF graph.
* Inference services: we provide metadata expansions services based on interlinked vocabularies in combination with ontologies. This inference approach can be used with the semantic tagging for unstructured data as wells as the structured data approaches.
+ Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inferences. The query can be configured to work for various scenarios.
* Vocabulary crossovers are a data modeling approach using interlinks from an ontology to relate
+ different vocabularies from similar or the same domains, which cover different areas or aspects of the domain. By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand the descriptions by traversing the interlinked graph.
+ different versions of the same vocabulary to automatically translate between those versions.
* Metadata inference based on interlinked vocabularies is similar to crossovers, but intended to represent different levels of abstraction, so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go more into detail.

# Prospects and Recommendations
We provide a first prototype to implement our approach. Next steps are bringing it into the actual data space in practice and representing and integrating further data sets from different data space participants. Also, the ML should actually benefit regarding quality in practice, which still needs to be verified.

# Conclusion

>>>>

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

@@

author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    affiliation: 
     name: The Semantic Web Company
     country: Austria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliation: 
      name: Ontotext
      country: Bulgaria
acm-metadata:
  keywords:
    - dataspaces
    - semantic interoperability
    - semantic technologies
    - ontologies
    - vocabulary hub
    - oil and gas
    - renewable energy
    - refineries
    - windfarms
  final: true
  # acmart-options: sigconf, review
  # copyright-year: 2024
  # acm-year: 2024
  copyright: acmcopyright
  # doi: 
  # conference-acronym: 
  # conference-name: 
  # conference-date: 
  # conference-location: 
  # isbn:

@@

# Introduction
The European data economy heavily depends on the availability of data and the technological foundation to make use of it. 
Different industries and communities can benefit mutually by sharing data with each other, thereby supporting their digitization business goals.
Machine-learning systems heavily rely on high volumes of high-quality training data. 
Examples are the well-known mobility dataspace and use cases like energy communities for running simulations.
To cope with the challenge of data sharing, Dataspace approaches like the IDS RAM were introduced to not only solve the technical aspects of data sharing, 
but also to establish methods for data sovereignty so that dataspace participants can determine themselves how and when others can make use of their data.

For making use of shared data, interoperability is a crucial factor. 
While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability.
IDS RAM is firmly based on semantic metadata for all aspects and actors of a dataspace.
Furthermore, it can leverage semantic interoperability approaches through the Vocabulary Hub
that stores ontologies and semantic thesauri for data descriptions and can dereference URIs to provide semantic details of data assets.

In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub 
built on the integration of the GraphDB semantic graph database and PoolParty Semantic Suite. 
We not only cover the core Vocabulary Hub functionalities and responsibilities in a dataspace, 
but also extend the idea towards a Semantic Layer for dataspaces, 
which provides even more services to easily and semi-automatically manage semantic interoperability.
Thus we raise the role of the Vocabulary Hub to not only store semantic assets, 
but also to bind semantic descriptions dynamically to incoming data, 
therefore enabling richer discovery and semantic data integration between datasets.

# Projects and Use Cases
In the expanding data economy landscape supported by dataspaces, 
we are working on several projects which address practical data sharing needs and were we experience the importance of semantic interoperability to fulfill various use cases. 
Here we present 3 examples from the [DataBri-X](https://databri-x.eu/) and [UNDERPIN](https://underpinproject.eu/) projects that cover various application domains.

## Energy Dataspace
In this use case, we run simulations for energy communities, which predict the behavior of the energy grid. 
To provide precise predictions, there is the need of high amounts of example data. 
This data is of numeric type and different tabular data sources need to be semantically consolidated.

## Legal Dataspace
Analyzing documents in the legal domain is a highly challenging task because of the complexity of legal information. 
We provide recommender services for insights into corpora of legal documents
that identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.

## Manufacturing/Maintenance Dataspace
Operating refineries and wind farms involves high maintenance costs. 
Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business-critical. 
For establishing predictive maintenance, we need high amounts of high quality training data
that can be obtained by data consolidation across different refinery and wind farm measurement sources.

The UNDERPIN project addresses this problem by creating a dataspace for refineries, wind farms, and their equipment suppliers.
Extensive time series data is exchanged, but their format and meaning is not standardized.
Describing the meaning of these time series facilitates semantic interoperability between datasets from different providers.
This description includes measured variable, quantity kind and unit of measure,
which sensor took the measurements, what it is attached to, the detailed model and connectivity of that industrial equipment.

## The Problem
"Standard" dataspace solutions provide access to data, but do not typically address the data integration problem.
If a consumer wants to consume datasets from different providers in an integrated way, 
they need to harmonize the data to some common model, convert it, store it in a database, 
and potentially implement entity linking (correlation of different records that are about the same real-world thing) and data fusion.

Above we gave just 3 examples to illustrate the heterogeneity of data in the context of different use cases. 
We aim to develop solutions that not only fulfill the dataspace needs, 
but also improve the situation regarding semantic interoperability by providing services that can be easily leveraged to implement such use cases.

# Approach and IT-Solution
To tackle this problem, we introduce Semantic Web standards and technologies to model and process data. 
Vocabularies and ontologies (expressed using RDFS, OWL and SKOS) can represent data with a clear semantics based on standards. 
We leverage GraphDB and PoolParty as two software components that implement these standards to process the data.

## IDS Vocabulary Hub
IDSA defines the architecture of dataspaces in the reference architecture model IDS RAM. 
One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services. 
These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
and thereby provides semantic interoperability in a standardized and machine-readable way.

## GraphDB Graph Database
TODO

## PoolParty Semantic Suite
PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.

@@

PoolParty is well suited to form the basis of a vocabulary management system within a dataspace, 
effectively implementing not only the Vocabulary Hub role, 
but also providing additional services that improve the expressiveness of common vocabularies and automate semantic interoperability.

@@

Both GraphDb and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL. They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.

# A Semantic Solution and its Benefits
When harmonizing data, we need to provide a method to automatically map and process it,
so this can be done efficiently on large datasets and/or regular updates in dataspaces. 
The benefit provided should support the ML training in such a way that the consolidated dataset is exposed as uniform training dataset.

## Semantic Layer Approach
The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
We aim to go one step further and envision a Semantic Layer for dataspaces, 
which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.

In the following, we discuss 
i) the services provided for the Vocabulary Hub component and 
ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.

Vocabulary Hub Services:
- RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
- Scalable vocabulary storage: cluster implementation for higher throughput and a fault tolerant architecture.
- Knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. 
  Vocabularies can be created and edited based on standards like RDFS, OWL and SKOS, and existing vocabularies can be reused and combined to fit specific dataspace needs.
- SPARQL service for semantic lookups: dereference URIs describing data assets to retrieve a semantic description of the entity from the graph database.

Semantic Layer Services support both unstructured and structured data assets:
- Unstructured data: semantic annotation services for documents based on concept tagging. 
  PoolParty’s KG-based NLP primarily uses SKOS taxonomies to identify concepts in documents, 
  but can be extended with OWL ontology elements (classes, properties) for increased expressiveness.
- Structured data: we support structured data harmonization methods, including semantic mapping and linking SKOS concepts
  - Ontology-Based Data Access (OBDA) is used for virtualization of relational data, i.e. accessing relational data as semantic data.
    It is well suited for semantic querying of large amounts of data, while keeping the original relational data in place (NoETL approach). 
    Ontologies are leveraged to construct the OBDA mapping; 
    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping. 
    Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
  - RDB to RDF Mapping Language (R2RML) [@souripriyadasR2RMLRDBRDF2012]
    and RDF Mapping Language (RML) [@RMLIntroduction2024]
    are intended for data transformation processes within a dataspace, 
    where the original source is used as a basis to create an RDF representation (ETL approach).
  - GraphDB supports both OBDA and R2RML through its integration of ONTOP [@bagosiOntopFrameworkOntology2014].
    The same mappings can be used for NoETL virtual data access, or for ETL (to materialize the semantic data).
  - CSV on the Web (CSVW) [@swirrlCSVWStandards2024] is a set of W3C standards 
    that allow detailed semantic description of tabular data using a JSON "manifest" file.
  - SKOS concepts are used for structured data in several ways. 
    Concepts can be directly added to RDF data via semantic links (properties from an ontology), 
    they can be used to annotate specific literals of an RDF graph,
    or they can be used to associate documents with (specific parts of) an RDF graph.
- Inference services: we provide metadata expansion services based on interlinked vocabularies in combination with ontologies. 
  This inference approach can be used with semantic tagging for unstructured data, as well as the structured data approaches.
  - Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inference. 
    The query can be configured to work for various scenarios.
- Vocabulary crosswalks are a data modeling approach using interlinks to relate:
  - Different vocabularies from similar or the same domain that cover different areas or aspects of the domain. 
    By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand queries and annotations by traversing the interlinked graph.
  - Different versions of the same vocabulary to automatically translate between those versions.
- Metadata inference based on interlinked vocabularies is similar to crosswalks, 
  but intended to represent different levels of abstraction, 
  so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. 
  For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go into more detail.

## Architecture and Implementation

# Conclusion and Future Work
We presented an approach for dynamically binding dataspace datasets
We have implemented a first prototype of our approach.

The next steps include:
- Bringing it into an actual dataspace in practice, representing and integrating datasets from different dataspace participants.
- Implementing use cases for richer discoverability and harmonized querying.
Also, the ML should actually benefit regarding quality in practice, which still needs to be verified.

# Acknowledgements
This work is partially supported by the Digital Europe programme project [UNDERPIN](https://underpinproject.eu/) (grant agreement 101123179)
